A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Additional Methods

### 2018-11-28 MS.tex
Major reformatting following meeting with Ace

\usepackage{xr}
\externaldocument{figures}
\externaldocument[S-]{supplementary_materials}

@@

@@

Data from MRR experiments in the Guerra et al. (2014) database were
examined and those with fewer than six recaptures and species with only
a single MRR study were excluded for the hierarchical analysis. Of the
232 data sets, 179 involved only females, 35 males, and 18 both sex
releases. For 102 data sets the age of the released mosquitoes was known
(the average age of released mosquitoes was 4.0 days) while in the other
cases it was unknown or unrecorded; in these cases we assumed the
mosquitoes were newly emerged at the time of release and return to this
assumption later. See Table \ref{S-tab:mrr_aggregateData} for a summary of other data characteristics.

@@

To access whether temperature is associated with LBL we used weather
records to calculate average temperatures at the MRR sites (see
Methods). Using both linear and quadratic temperature terms in
regressions, we found no significant relationship between study-site
temperature and LBL (overall or within genus) for the 238 datasets we
analysed (Fig. \ref{fig:mrr_temperature}). This result held if, instead of pooling results from
all time series, we considered the four species with the most data
individually (\emph{Ae. aegypti}, \emph{Cx. tarsalis}, \emph{A. gambiae s.l.} and
\emph{A. culicifacies s.l.}; Fig. \ref{S-fig:mrr_ThreeSpeciesVersusTemperature}).

@@

Dissection allows the number of completed gonotrophic cycles to be
counted and from this the mean number of cycles before death was
estimated. Overall, the mean number of cycles completed in a lifetime
was 1.3 (posterior mean; Fig. \ref{fig:dissection_lifetimes_exponential}; Table \ref{S-tab:dissection_estimated_lifespans}) and across the 131 studies, 95\% of
the individual time series estimates were less than 3 (Fig. \ref{fig:dissection_individualEstimates_allSpecies}). The
estimated greatest number of cycles was for \emph{Anopheles sergentii}
(3.0 cycles; posterior mean) which is adapted to desert conditions (it
is known as the ``oasis vector'' of malaria) and may have evolved
greater longevity. The important African malaria vector \emph{A.
gambiae s.l.} was estimated to be the second longest living (2.4 cycles;
posterior mean). The smallest estimated mean number of gonotrophic
cycles was for \emph{Anopheles} \emph{bellator} (0.6 cycles; posterior
mean) which transmits malaria in Brazil's Atlantic Forest. There were
significant differences in estimated lifetime gonotrophic cycles amongst
species (ANOVA: \emph{F}\textsubscript{24,106} =2.2, \emph{p}
\textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{24}\), \emph{p}\textless{}0.01).

@@

To compare the two methods, we converted numbers of gonotrophic cycle
(physiological age) into lifespan (chronological age) as described in
the SOM using estimates of gonotrophic cycle duration (Fig. \ref{S-fig:dissection_gonotrophicCycleRaw}). Table \ref{S-tab:dissection_estimated_lifespans_chron} provides posterior summaries of chronological for the species and genera in the dissection dataset (see also Fig. \ref{S-fig:dissection_lifetimes_exponential_chron}). For ten species, we had enough data from both species to make a comparison, and there was a positive correlation (not
statistically significant; Pearson correlation $\rho=0.42$, $n=10$, $p=0.23$)
between the two measures (Fig. \ref{fig:comparison}), and in only one case -- for \emph{A.
darlingi} - there was a significant difference in the
time-series level LBLs (Table \ref{S-tab:comparison}).

@@

We conducted a power analysis of MRR experiments to determine
whether typical experimental characteristics could detect senescence.
Here we calculated the power of a maximum likelihood estimator of the
`senescence parameter' $\beta$ of the Gompertz survival function (see Table
\ref{S-tab:mrr_survivalDescription}) for case study populations with three different levels of
senescence (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}A). This analysis indicated that power to detect
senescence strongly depends on study length (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}B) but is
insensitive to release size (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}C). Clements and Patterson (1981)
conducted a meta-analysis of MRR and dissection-based field experiments
and found evidence of an increasing risk of mortality hazard with age
that is similar in magnitude to that of the `mild' case considered
above. For this case, detecting senescence with a power of 80\% requires
a study length of at least 18 days. Since the median study duration for
experiments included in our analysis was 10 days (Table \ref{S-tab:mrr_IndividualData}) this could
partly explain our failure to detect senescence at the species level. A
number of experiments have found evidence of age-dependence in
laboratory populations \citep{styer2007mosquitoes,dawes2009anopheles}.
However, the artificially benign environment of the laboratory means
mosquitoes live considerably longer than in the wild, where they may die
because of exogenous factors, before the effects of physiological
decline have had time to manifest. Field experiments have also found
evidence for age-dependent mortality. Harrington et al. (2008) conducted
a field experiment where mosquitoes reared under laboratory conditions
were marked and released at different ages. Analysis of the resultant
MRR time-series indicated that mosquito mortality increases with age at
release. It is possible, however, that this field experiment suffers
from the same biases as laboratory-based approaches, because the
released mosquitoes were often of ages considerably higher (up to 20
days) than typical estimates of wild mosquito lifespan.

As ethical concerns of contributing to disease burden are more often
considered, it is now less common for MRR experiments to release female
mosquitoes versus males than historically (Fig. \ref{S-fig:mrr_sexReleasesOverTime}). Our analysis
indicates that females outlive male mosquitoes by approximately 1.2 days
(Fig. \ref{fig:mrr_sexDifferences_without_sugar_nor_blood}), meaning that differences between the sexes may exist for
other ecological parameters determinable by MRR. This suggests that
continued field entomological work on contained releases of mosquitoes
in semi-field sites or large microcosms may be a valuable source of
information on female mosquito ecology.

@@

There is evidence mainly from laboratory studies that temperature
modulates mosquito ecology and behaviour \citep{yang2009assessing,brady2013modelling,murdock2012complex,beck2013effect}. The
locations and times of year over which the MRR studies were conducted
encompassed a large range of average air temperatures, from
approximately 10 \textsuperscript{o}C to 35 \textsuperscript{o}C and,
within this, we determined no relationship between lifespan and
temperature across all time series (Fig. \ref{fig:mrr_temperature}) or, for any of the species
with the most data (Fig. \ref{S-fig:mrr_ThreeSpeciesVersusTemperature}). It is possible that by considering a raw
average of air temperature across the month, this ignored, more complex, interactions between temperature and lifespan. It is also possible that by ignoring the effects of
rainfall (the historical data on rainfall is less likely to be reliable
for a given location), that this masked a more complex interaction
between longevity and temperature. The observed laboratory relationship
between lifespan and temperature, however, may not be as robust in the
field if mosquitoes adjust their behaviours (such as, by seeking shade)
in reaction to changes in temperature. More work exploring the
relationship between mosquito ecology and temperature in semi-field
experiments may be useful in probing these interactions further.

@@

>>>>

\usepackage{xr}
\externaldocument{figures}
\externaldocument[S-]{supplementary_materials}
\externaldocument[SM-]{supplementary_materials_methods}

@@

\section{Author summary}

\section{Author contributions}
HCJG, AN and BL were involved in conceptualising this study. BL was responsible for data curation and the formal analysis of the data. BL and AN developed the statistical methodology and conducted the investigation. All authors were involved in drafting the original manuscript and revising it.

@@

@@

To access whether temperature is associated with LBL we used weather
records to calculate average temperatures at the MRR sites (see
Methods). Using both linear and quadratic temperature terms in
regressions, we found no significant relationship between study-site
temperature and LBL (overall or within genus) for the 238 datasets we
analysed (Fig. \ref{S-fig:mrr_temperature}). This result held if, instead of pooling results from
all time series, we considered the four species with the most data
individually (\emph{Ae. aegypti}, \emph{Cx. tarsalis}, \emph{A. gambiae s.l.} and
\emph{A. culicifacies s.l.}; Fig. \ref{S-fig:mrr_ThreeSpeciesVersusTemperature}).

@@

Dissection allows the number of completed gonotrophic cycles to be
counted and from this the mean number of cycles before death was
estimated. Overall, the mean number of cycles completed in a lifetime
was 1.3 (posterior mean; Fig. \ref{fig:dissection_lifetimes_exponential}; Table \ref{S-tab:dissection_estimated_lifespans}) and across the 131 studies, 95\% of
the individual time series estimates were less than 3 (Fig. \ref{S-fig:dissection_individualEstimates_allSpecies}). The
estimated greatest number of cycles was for \emph{Anopheles sergentii}
(3.0 cycles; posterior mean) which is adapted to desert conditions (it
is known as the ``oasis vector'' of malaria) and may have evolved
greater longevity. The important African malaria vector \emph{A.
gambiae s.l.} was estimated to be the second longest living (2.4 cycles;
posterior mean). The smallest estimated mean number of gonotrophic
cycles was for \emph{Anopheles} \emph{bellator} (0.6 cycles; posterior
mean) which transmits malaria in Brazil's Atlantic Forest. There were
significant differences in estimated lifetime gonotrophic cycles amongst
species (ANOVA: \emph{F}\textsubscript{24,106} =2.2, \emph{p}
\textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{24}\), \emph{p}\textless{}0.01).

@@

To compare the two methods, we converted numbers of gonotrophic cycles
(physiological age) into lifespan (chronological age) as described in
the SOM using estimates of gonotrophic cycle duration (Fig. \ref{S-fig:dissection_gonotrophicCycleRaw}). Table \ref{S-tab:dissection_estimated_lifespans_chron} provides posterior summaries of chronological for the species and genera in the dissection dataset (see also Fig. \ref{S-fig:dissection_lifetimes_exponential_chron}). For ten species, we had enough data from both species to make a comparison, and there was a positive correlation (not
statistically significant; Pearson correlation $\rho=0.42$, $n=10$, $p=0.23$)
between the two measures (Fig. \ref{fig:comparison}), and in only one case -- for \emph{A.
darlingi} - there was a significant difference in the
time-series level LBLs (Table \ref{S-tab:comparison}).

@@

We conducted a power analysis of MRR experiments to determine
whether typical experimental characteristics could detect senescence.
Here we calculated the power of a maximum likelihood estimator of the
`senescence parameter' $\beta$ of the Gompertz survival function (see Table
\ref{SM-tab:mrr_survivalDescription}) for case study populations with three different levels of
senescence (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}A). This analysis indicated that power to detect
senescence strongly depends on study length (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}B) but is
insensitive to release size (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}C). Clements and Patterson (1981)
conducted a meta-analysis of MRR and dissection-based field experiments
and found evidence of an increasing risk of mortality hazard with age
that is similar in magnitude to that of the `mild' case considered
above. For this case, detecting senescence with a power of 80\% requires
a study length of at least 18 days. Since the median study duration for
experiments included in our analysis was 10 days (Table \ref{SM-tab:mrr_IndividualData}) this could
partly explain our failure to detect senescence at the species level. A
number of experiments have found evidence of age-dependence in
laboratory populations \citep{styer2007mosquitoes,dawes2009anopheles}.
However, the artificially benign environment of the laboratory means
mosquitoes live considerably longer than in the wild, where they may die
because of exogenous factors, before the effects of physiological
decline have had time to manifest. Field experiments have also found
evidence for age-dependent mortality. Harrington et al. (2008) conducted
a field experiment where mosquitoes reared under laboratory conditions
were marked and released at different ages. Analysis of the resultant
MRR time-series indicated that mosquito mortality increases with age at
release. It is possible, however, that this field experiment suffers
from the same biases as laboratory-based approaches, because the
released mosquitoes were often of ages considerably higher (up to 20
days) than typical estimates of wild mosquito lifespan.

As ethical concerns of contributing to disease burden are more often
considered, it is now less common for MRR experiments to release female
mosquitoes versus males than historically (Fig. \ref{SM-fig:mrr_sexReleasesOverTime}). Our analysis
indicates that females outlive male mosquitoes by approximately 1.2 days
(Fig. \ref{fig:mrr_sexDifferences_without_sugar_nor_blood}), meaning that differences between the sexes may exist for
other ecological parameters determinable by MRR. This suggests that
continued field entomological work on contained releases of mosquitoes
in semi-field sites or large microcosms may be a valuable source of
information on female mosquito ecology.

@@

There is evidence mainly from laboratory studies that temperature
modulates mosquito ecology and behaviour \citep{yang2009assessing,brady2013modelling,murdock2012complex,beck2013effect}. The
locations and times of year over which the MRR studies were conducted
encompassed a large range of average air temperatures, from
approximately 10 \textsuperscript{o}C to 35 \textsuperscript{o}C and,
within this, we determined no relationship between lifespan and
temperature across all time series (Fig. \ref{S-fig:mrr_temperature}) or, for any of the species
with the most data (Fig. \ref{S-fig:mrr_ThreeSpeciesVersusTemperature}). It is possible that by considering a raw
average of air temperature across the month, this ignored, more complex, interactions between temperature and lifespan. It is also possible that by ignoring the effects of
rainfall (the historical data on rainfall is less likely to be reliable
for a given location), that this masked a more complex interaction
between longevity and temperature. The observed laboratory relationship
between lifespan and temperature, however, may not be as robust in the
field if mosquitoes adjust their behaviours (such as, by seeking shade)
in reaction to changes in temperature. More work exploring the
relationship between mosquito ecology and temperature in semi-field
experiments may be useful in probing these interactions further.

@@

Data from MRR experiments in the Guerra et al. (2014) database were
examined and those with fewer than six recaptures and species with only
a single MRR study were excluded for the hierarchical analysis. Of the
232 data sets, 179 involved only females, 35 males, and 18 both sex
releases. For 102 data sets the age of the released mosquitoes was known
(the average age of released mosquitoes was 4.0 days) while in the other
cases it was unknown or unrecorded; in these cases we assumed the
mosquitoes were newly emerged at the time of release and return to this
assumption later. See Table \ref{SM-tab:mrr_aggregateData} for a summary of other data characteristics.
